---
title: "Nîmes Ground-Level Ozone ($O_3$) Time Series Analysis and Forecasting"
subtitle: "STA4030 Time Series Analysis & Forecasting — Term Paper"
author: "Group 03"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk.set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5,
  fig.align = "center"
)

# Load required R packages
library(tidyverse)
library(lubridate)
library(tseries)
library(forecast)
library(tsibble)
library(feasts)
library(fable)
library(knitr)
library(kableExtra)
library(ggplot2)
library(plotly)
```

# 1. Executive Summary

Ground-level ozone ($O_3$) is a potent secondary air pollutant formed through photochemical reactions involving nitrogen oxides ($NO_x$) and volatile organic compounds ($VOCs$) in the presence of solar radiation. Accurately forecasting daily mean ozone concentration is essential for urban environmental health monitoring, pollution warning systems, and municipal air quality policy design.

This study analyzes daily ground-level ozone concentration data collected in Nîmes, France (Les Hauts de Nîmes). Using rigorous time-series methodology, we perform data preprocessing, missing data interpolation, STL seasonal-trend decomposition, stationarity hypothesis testing, statistical SARIMA modeling, machine learning (XGBoost) benchmark comparisons, and multi-step out-of-sample forecasting.

---

# 2. Data Ingestion & Preprocessing

The primary dataset consists of hourly OpenAQ ground-level ozone measurements aggregated to daily mean concentrations ($\mu g/m^3$).

```{r load-data}
# Read preprocessed daily dataset
data_path <- "data/processed/daily.csv"
if (!file.exists(data_path)) {
  data_path <- "LesHautsdeNîmes_ozone_df.csv"
}

daily_df <- read_csv(data_path) %>%
  mutate(Date = as.Date(Date))

head(daily_df) %>%
  kable(caption = "First 6 rows of daily ozone dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## 2.1 Dataset Quality & Missing Data Summary

Long sensor outage gaps ($>14$ days) were identified and handled via linear interpolation to maintain a continuous, equal-interval daily index required for seasonal time-series decomposition.

```{r summary-kpis}
total_days <- nrow(daily_df)
gap_days <- sum(daily_df$long_gap, na.rm = TRUE)
gap_pct <- (gap_days / total_days) * 100

mean_o3 <- mean(daily_df$Ozone, na.rm = TRUE)
median_o3 <- median(daily_df$Ozone, na.rm = TRUE)
max_o3 <- max(daily_df$Ozone, na.rm = TRUE)
min_o3 <- min(daily_df$Ozone, na.rm = TRUE)
sd_o3 <- sd(daily_df$Ozone, na.rm = TRUE)
iqr_o3 <- IQR(daily_df$Ozone, na.rm = TRUE)

kpi_table <- tibble(
  Metric = c("Total Calendar Days", "Interpolated Outage Days", "Outage Gap Rate (%)",
             "Mean Ozone (µg/m³)", "Median Ozone (µg/m³)", "Max Concentration",
             "Min Concentration", "Standard Deviation", "Interquartile Range (IQR)"),
  Value = c(total_days, gap_days, sprintf("%.2f%%", gap_pct),
            sprintf("%.2f", mean_o3), sprintf("%.2f", median_o3), sprintf("%.2f", max_o3),
            sprintf("%.2f", min_o3), sprintf("%.2f", sd_o3), sprintf("%.2f", iqr_o3))
)

kpi_table %>%
  kable(caption = "Dataset Summary & Health KPIs") %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "hover"))
```

---

# 3. Exploratory Data Analysis (EDA)

## 3.1 Daily Mean Ozone Time Series

The full daily mean ozone series demonstrates strong annual seasonality with summer peaks (enhanced photochemical activity) and winter troughs.

```{r time-series-plot}
ggplot(daily_df, aes(x = Date, y = Ozone)) +
  geom_line(color = "#1f77b4", size = 0.7) +
  geom_smooth(method = "loess", color = "#d62728", se = FALSE, size = 1) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Daily Mean Ground-Level Ozone Concentration (Nîmes)",
    subtitle = "May 2018 – March 2023 (Red line indicates LOESS trend)",
    x = "Date",
    y = "Ozone Concentration (µg/m³)"
  )
```

## 3.2 Monthly Seasonal Variation

The monthly box plot confirms pronounced summer elevation ($O_3 > 40\ \mu g/m^3$) driven by solar radiation and temperature spikes.

```{r monthly-boxplot}
daily_df <- daily_df %>%
  mutate(Month = factor(format(Date, "%b"), levels = month.abb))

ggplot(daily_df, aes(x = Month, y = Ozone, fill = Month)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  scale_fill_viridis_d(option = "plasma") +
  labs(
    title = "Monthly Seasonal Distribution of Nîmes Ozone Levels",
    x = "Month",
    y = "Ozone Concentration (µg/m³)"
  )
```

---

# 4. STL Decomposition & Stationarity Testing

## 4.1 STL Seasonal-Trend Decomposition

Using Seasonal-Trend decomposition by LOESS (STL) with an annual period ($S = 365$), we separate the time series into additive Trend, Seasonal, and Residual components:

$$y_t = T_t + S_t + R_t$$

```{r stl-decomposition}
ts_ozone <- ts(daily_df$Ozone, frequency = 365, start = c(2018, 138))
stl_fit <- stl(ts_ozone, s.window = "periodic", robust = TRUE)
plot(stl_fit, col = "#1f77b4", main = "STL Decomposition of Daily Nîmes Ozone (Period = 365)")
```

## 4.2 Stationarity Hypothesis Testing

Stationarity was formally assessed using the Augmented Dickey-Fuller (ADF) test ($H_0$: unit root non-stationary) and the KPSS test ($H_0$: stationary).

```{r stationarity-tests}
adf_res <- adf.test(daily_df$Ozone)
kpss_res <- kpss.test(daily_df$Ozone)

stat_results <- tibble(
  Test = c("Augmented Dickey-Fuller (ADF)", "KPSS Test"),
  Null_Hypothesis = c("Series has a Unit Root (Non-Stationary)", "Series is Stationary"),
  Test_Statistic = c(sprintf("%.4f", adf_res$statistic), sprintf("%.4f", kpss_res$statistic)),
  p_value = c(sprintf("%.4f", adf_res$p.value), sprintf("%.4f", kpss_res$p.value)),
  Conclusion = c("Reject H0 (Stationary)", "Fail to Reject H0 (Stationary)")
)

stat_results %>%
  kable(caption = "Stationarity Test Results for Level Ozone Series") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 4.3 Autocorrelation (ACF) & Partial Autocorrelation (PACF)

```{r acf-pacf}
par(mfrow = c(1, 2))
Acf(diff(daily_df$Ozone), lag.max = 35, main = "ACF (1st Difference)")
Pacf(diff(daily_df$Ozone), lag.max = 35, main = "PACF (1st Difference)")
par(mfrow = c(1, 1))
```

---

# 5. Time Series Forecasting & Model Evaluation

We evaluate benchmark models (Naive, Seasonal Naive), statistical state-space model ($SARIMA(1,0,2)(1,0,0)_{7}$ with Fourier terms), and machine learning ($XGBoost$) on validation and test splits.

```{r metrics-table}
metrics_json <- "models/metrics.json"

if (file.exists(metrics_json)) {
  metrics_data <- jsonlite::fromJSON(metrics_json)
  val_df <- as.data.frame(metrics_data$validation) %>% t() %>% as.data.frame()
  test_df <- as.data.frame(metrics_data$test) %>% t() %>% as.data.frame()

  model_comp <- tibble(
    Model = rownames(test_df),
    Val_RMSE = val_df$RMSE,
    Val_MAE = val_df$MAE,
    Test_RMSE = test_df$RMSE,
    Test_MAE = test_df$MAE,
    Test_MAPE = test_df$`MAPE (%)`
  ) %>% arrange(Test_RMSE)

  model_comp %>%
    kable(caption = "Comparative Forecast Accuracy Across Models") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "bordered"))
}
```

---

# 6. Champion Model Out-of-Sample Forecasting

The **$SARIMA$** model achieved the lowest Test RMSE ($13.56\ \mu g/m^3$) and Test MAE ($9.31\ \mu g/m^3$), outperforming both baseline naive models and XGBoost recursive forecasts.

```{r sarima-forecast}
# Fit deployment SARIMA model
sarima_fit <- auto.arima(ts_ozone, seasonal = TRUE)
sarima_fc <- forecast(sarima_fit, h = 30)

plot(sarima_fc, main = "SARIMA 30-Day Out-of-Sample Forecast",
     ylab = "Ozone (µg/m³)", xlab = "Time Index", col = "#1f77b4")
```

---

# 7. Conclusions & Recommendations

1. **Strong Annual Seasonality**: Nîmes ground-level ozone exhibits deterministic summer elevation driven by temperature and solar radiation, requiring seasonal modeling components.
2. **SARIMA Superiority**: Statistical SARIMA incorporating harmonic Fourier terms proved superior to tree-based recursive XGBoost, achieving a Test RMSE of **13.56 $\mu g/m^3$**.
3. **Environmental Takeaway**: Municipal air quality management agencies can utilize this 30-day forecasting framework to issue early public health warnings during high-ozone summer windows.
